---
title: "Predicting Life Expectancy in U.S. Cities: How A Machine Learning Solution May Aid In Budgetary Decision-Making"
subtitle: "Compiled Code"
author: Nikolas Tubert
date: '2025-12-04'
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: true
---

# LOAD LIBRARIES
```{r, warning = FALSE, message = FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(mice)
library(ggformula)
library(ggplot2)
library(scales)
library(knitr)
library(gridExtra)
library(broom)
library(naniar)
library(caret)
library(MASS)
library(ggforce)
library(corrplot)
library(car)
library(packcircles)
library(kableExtra)
library(grid)
library(xgboost)
library(glmnet)
library(nnet)
library(SHAPforxgboost)
library(pdp)
```

# DATA COLLECTION, CLEANING AND PREPROCESSING
In this section of code, I load in the data from a .csv file, explore the overall data structure, perform some initial data cleaning steps, impute missing values, and assess the presence of outliers.

## Read-in Data
```{r, include=FALSE}
bchc <- read_csv("BigCitiesHealth.csv")
```

## Data Structure
```{r}
dim(bchc)
summary(bchc)
```

## Heatmap of real vs. proxy values
```{r}
# Count rows by the two proxy variables
proxy_counts <- bchc %>% count(geo_label_proxy_or_real, date_label_proxy_or_real)

# Heatmap theme
heatmap_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 13, hjust = 0.5),
    axis.text.x = element_text(hjust = 1),
    panel.grid = element_blank()
  )

# Heatmap
ggplot(proxy_counts, aes(x = geo_label_proxy_or_real, y = date_label_proxy_or_real, fill = n)) +
  geom_tile(color = "grey80") +
  geom_text(aes(label = comma(n)), color = "#c77904", size = 4, fontface="bold") +
  scale_fill_gradientn(
    colors = c("#f7fbff", "#6baed6", "#08306b"),
    values = rescale(c(0, max(proxy_counts$n) / 3, max(proxy_counts$n))),
    na.value = "white",
    labels = comma
  ) +
  labs(
    title = "Heatmap of Row Counts",
    subtitle = "By Proxy Indicators",
    x = "Geographic Proxy Indicator",
    y = "Date Proxy Indicator",
    fill = "Row Count"
  ) +
  heatmap_theme
```

## Initial Data Cleaning
```{r}
# Select, rename, and clean columns
bchc <- bchc %>%
  dplyr::select(
    metric_item_label,
    geo_label_citystate,
    date_label,
    value,
    strata_race_label,
    strata_sex_label
  ) %>%
  rename(
    metric = metric_item_label,
    city = geo_label_citystate,
    year = date_label,
    race = strata_race_label,
    sex = strata_sex_label
  ) %>%
  mutate(
    metric = tolower(gsub("[,().-/]", "", gsub(" ", "_", metric))),
    city = factor(city),
    year = factor(year),
    race = factor(race),
    sex = factor(sex)
  ) %>%
  filter(city != "U.S. Total")

# Reshape data frame
bchc_wide <- bchc %>%
  pivot_wider(
    names_from = metric,
    values_from = value
  )

# Correct any invalid column names
colnames(bchc_wide) <- make.names(colnames(bchc_wide), unique = TRUE)

# Sort values
bchc_wide <- arrange(bchc_wide, city, race, sex, year)
```

## Inspect Missingness, Drop Data, and Impute
### Inital Assessment of Missingness
```{r}
na_cells <- sum(is.na(bchc_wide))
total_cells <- sum(!is.na(bchc_wide)) + sum(is.na(bchc_wide))
print(paste0("Original Missing Percent: ", na_cells / total_cells))
```

### Drop Data
#### Step 1: Drop rows with high % of info missing (retaining rows for Long Beach, CA in year 2023)
```{r}
row_missing_pct <- rowMeans(is.na(bchc_wide))
missing_pct_p75 <- quantile(row_missing_pct, 0.75)
bchc_wide <- bchc_wide[row_missing_pct <= missing_pct_p75 | (bchc_wide$city == "Long Beach, CA" & bchc_wide$year == "2023"), ]
```

#### Step 2: Drop metrics with high % info missing
```{r}
# Summary of column missingness
total_rows <- nrow(bchc_wide)
na_by_metric <- data.frame(
  metric = names(bchc_wide),
  populated_rows = sapply(bchc_wide, function(x) sum(!is.na(x))),
  na_rows = sapply(bchc_wide, function(x) sum(is.na(x))),
  na_pct = sapply(bchc_wide, function(x) sum(is.na(x)) / total_rows),
  row.names = NULL
) %>%
  filter(
    metric != "city",
    metric != "year",
    metric != "race",
    metric != "sex",
  ) %>%
  arrange(populated_rows)

# Histogram theme
hist_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 13, hjust = 0.5)
  )

# Histogram
ggplot(na_by_metric, aes(x = na_pct)) +
  geom_histogram(fill="#005AB5", color="#c9c9c9", binwidth=0.03) +
  geom_vline(xintercept=0.75, linetype = "dashed", color="#b30404", size=1) +
  labs(title = "Histogram of Variable Missingness",
       subtitle = "Count of Variables by Missing Percent",
       x = "Missing Percent",
       y = "Count of Variables") +
  hist_theme
```

```{r}
# Drop columns with high missing %
metrics_to_drop <- na_by_metric %>%
  filter(na_pct > 0.75) %>%
  pull(metric)

bchc_wide <- bchc_wide %>%
  dplyr::select(-all_of(metrics_to_drop))
```

### Final Assessment of Missingness
```{r}
na_cells <- sum(is.na(bchc_wide))
total_cells <- sum(!is.na(bchc_wide)) + sum(is.na(bchc_wide))
print(paste0("Missing Percent: ", na_cells / total_cells))
```

### Impute
#### Step 1: Test for MAR
```{r, warning=FALSE}
# Create Boolean columns for each variable (TRUE if NA, FALSE if populated)
vars_to_check <- names(bchc_wide %>% dplyr::select(-c(city, year, race, sex)))

missing_inds <- bchc_wide %>%
  mutate(across(all_of(vars_to_check), ~is.na(.), .names = "missing_{.col}"))

# For each Boolean column, run logistic regression with city, year, race, and sex as independent variables
results <- list()

for (var in vars_to_check) {
  formula_str <- paste0("missing_", var, " ~ city + year + race + sex")
  formula_obj <- as.formula(formula_str)
  
  model <- glm(formula_obj, data = missing_inds, family = binomial)
  results[[var]] <- model
}

# Summarize p-values
pval_summary <- lapply(names(results), function(var) {
  model <- results[[var]]
  if (!is.null(model)) {
    tidy_model <- broom::tidy(model)
    tidy_model$variable <- var
    return(tidy_model)
  }
})

pval_df <- do.call(rbind, pval_summary)

# Distinguish between variables that have significant terms (p-value < 0.05) and variables with no significant terms
mar_flags <- pval_df %>%
  filter(term != "(Intercept)", p.value < 0.05) %>%
  distinct(variable) %>%
  mutate(missingness_type = "MAR")

all_vars <- data.frame(variable = vars_to_check)
missingness_types <- all_vars %>%
  left_join(mar_flags, by = "variable") %>%
  mutate(missingness_type = ifelse(is.na(missingness_type), "MCAR or MNAR", missingness_type))

print(paste(nrow(mar_flags), " variables of the total ",  nrow(missingness_types), "have significant terms in the logistic regression models"))
```

#### Step 2: Imputation using Predictive Mean Matching
```{r, warning=FALSE}
# Initiate MICE method
mice_method <- make.method(bchc_wide)
mice_method[] <- ""
mice_method[vars_to_check] <- "pmm"

# Construct predictor matrix
mice_pred <- make.predictorMatrix(bchc_wide)
mice_pred[,] <- 0
mice_pred[vars_to_check, c("city", "year", "race", "sex")] <- 1
mice_pred[vars_to_check, vars_to_check] <- 1

# Impute
mice <- mice(bchc_wide,
  m = 1,
  method = mice_method,
  predictorMatrix = mice_pred,
  seed = 10,
  printFlag = FALSE)

bchc_imputed <- complete(mice)
```

## Outlier Detection
### Step 1: Find all variables that have outliers
```{r}
outlier_detect <- function(x) {
  q1 <- quantile(x, 0.25)
  q3 <- quantile(x, 0.75)
  iqr <- q3 - q1
  lower <- q1 - 1.5 * iqr
  upper <- q3 + 1.5 * iqr
  any(x < lower | x > upper)
}

vars_with_outliers <- Filter(function(var) outlier_detect(bchc_imputed[[var]]), vars_to_check)
```

### Step 2: Example of transformation/scaling to be used
```{r}
var <- "premature_death"

# Calculate scaled values
bchc_scaled <- dplyr::select(bchc_imputed, all_of(var))
pre_yeo <- preProcess(bchc_scaled[, var, drop = FALSE], method = c("YeoJohnson", "center", "scale"))
bchc_scaled[[paste0(var, "_yeo")]] <- predict(pre_yeo, bchc_scaled[, var, drop = FALSE])[, 1]

pre_range <- preProcess(bchc_scaled[, var, drop = FALSE], method = c("YeoJohnson", "range"))
bchc_scaled[[paste0(var, "_range")]] <- predict(pre_range, bchc_scaled[, var, drop = FALSE])[, 1]

# Count outliers
count_outliers <- function(x) {
  stats <- boxplot.stats(x)
  length(stats$out)
}

n_outliers <- count_outliers(bchc_scaled[[var]])
n_outliers_yeo <- count_outliers(bchc_scaled[[paste0(var, "_yeo")]])
n_outliers_range <- count_outliers(bchc_scaled[[paste0(var, "_range")]])

# Boxplot theme
boxplot_theme <- theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18, color = "#2c3e50", hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 14, color = "#34495e", hjust = 0.5),
    axis.title.y = element_text(size = 14, face = "bold", color = "#2c3e50"),
    axis.text.y = element_text(size = 12, color = "#2c3e50"),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank()
  )

# Raw boxplot
ggplot(bchc_scaled, aes(y = .data[[var]])) +
  geom_boxplot(
    fill = "#3498db",                
    outlier.colour = "#b30404",      
    outlier.shape = 8,
    outlier.size = 2
  ) +
  annotate("text", x = 0.3, y = max(bchc_scaled[[var]]), 
           label = paste("Outliers:", n_outliers),
           hjust = 1.1, vjust = -0.5, color = "#b30404", size = 4) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = paste("Boxplot of", var),
       subtitle = "No Transformation or Scaling",
       y = paste(var)) +
  boxplot_theme

# Yeo-Johnson + Center/Scale boxplot
ggplot(bchc_scaled, aes(y = .data[[paste0(var, "_yeo")]])) +
  geom_boxplot(
    fill = "#2ecc71",                
    outlier.colour = "#b30404",
    outlier.shape = 8,
    outlier.size = 2
  ) +
  annotate("text", x = 0.3, y = max(bchc_scaled[[paste0(var, "_yeo")]]), 
           label = paste("Outliers:", n_outliers_yeo),
           hjust = 1.1, vjust = -0.5, color = "#b30404", size = 4) +
  labs(
    title = paste("Boxplot of", var),
    subtitle = "Yeo-Johnson + Traditional Scaling",
    y = paste(var, "\nYeo-Johnson + Scaling")
  ) +
  boxplot_theme

# Yeo-Johnson + Min-Max boxplot
ggplot(bchc_scaled, aes(y = .data[[paste0(var, "_range")]])) +
  geom_boxplot(
    fill = "#f1c40f",                
    outlier.colour = "#b30404",
    outlier.shape = 8,
    outlier.size = 2
  ) +
  annotate("text", x = 0.3, y = max(bchc_scaled[[paste0(var, "_range")]]), 
           label = paste("Outliers:", n_outliers_range),
           hjust = 1.1, vjust = -0.5, color = "#b30404", size = 4) +
  labs(
    title = paste("Boxplot of", var),
    subtitle = "Yeo-Johnson + Min-Max Scaling",
    y = paste(var, "\nYeo-Johnson + Min-Max Scaling")
  ) +
  boxplot_theme
```

# EXPLORATORY DATA ANALYSIS
In this section of code, I explore the remaining potential predictor variables, dropping redundant features based on correlation analysis and engineering one new feature. Additionally, I explore descriptive statistics related to the dependent variable and the independent variables, while also examing some relationships between predictors and the response variable.

## Assess Available Data
### List of columns
```{r}
colnames(bchc_imputed)
```

### Categorize Columns and Visualize with Bubble Map
```{r}
# Define categories
demographic <- c("city", "year", "race", "sex")
disease_mortality <- c("all_cancer_deaths", "lung_cancer_deaths", "colorectal_cancer_deaths", "breast_cancer_deaths", "prostate_cancer_deaths",
                       "cardiovascular_disease_deaths", "heart_disease_deaths", "diabetes_deaths", "pneumonia_or_influenza_deaths", "hiv.related_deaths")
other_mortality <- c("deaths_from_all_causes", "premature_death", "gun_deaths_firearms", "injury_deaths", "motor_vehicle_deaths", "police_killings",
                     "homicides", "infant_deaths", "drug_overdose_deaths", "opioid_overdose_deaths", "suicide")
healthcare <- c("uninsured_all_ages", "uninsured_child", "prenatal_care", "flu_vaccinations_medicare", "people_with_disabilities", 
                "hivaids_prevalence", "teen_births", "low_birthweight")
transportation <- c("walking_to_work", "public_transportation_use", "drives_alone_to_work")
socioeconomic <- c("single.parent_families", "owner_occupied_housing", "renters_vs_owners", "college_graduates", "poverty_in_all_ages",
                   "poverty_in_children", "per.capita_household_income", "unemployment", "service_workers")

var_groups <- list(
  Demographic = demographic,
  `Disease Mortality` = disease_mortality,
  `Other Mortality` = other_mortality,
  Healthcare = healthcare,
  Transportation = transportation,
  Socioeconomic = socioeconomic
)

# Create data frame of category variable
bubble_data <- bind_rows(
  lapply(names(var_groups), function(group) {
    data.frame(
      group = group,
      var_count = length(var_groups[[group]])
    )
  })
)
bubble_data$area <- bubble_data$var_count

# Build bubble map
packing <- circleProgressiveLayout(bubble_data$area, sizetype = "area")
bubble_data <- bind_cols(bubble_data, packing)

circle_polygons <- circleLayoutVertices(packing, npoints = 50)

ggplot() +
  geom_polygon(data = circle_polygons,
               aes(x, y, group = id, fill = as.factor(id)),
               color = "black", alpha = 0.4) +
  geom_text(data = bubble_data,
            aes(x, y + radius * 0.2, label = group),
            size = 3.5, fontface = "bold") +
  geom_text(data = bubble_data,
            aes(x, y - radius * 0.2, label = paste0(var_count, " variables")),
            size = 3) +
  coord_equal() +
  theme_void() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
        plot.subtitle = element_text(face = "italic", size = 13, hjust = 0.5)) +
  labs(title = "Predictor Variable Categories",
       subtitle = "Count of Predictors Per Category")

```

## Multicollinearity
### Build Correlation Matrix
```{r}
# Get correlation matrix of numeric predictors
num_predictors <- bchc_imputed[, setdiff(names(bchc_imputed), c("city", "year", "race", "sex", "life_expectancy"))]
corr_matrix <- cor(num_predictors, use = "pairwise.complete.obs")

# Find pairs of predictors that are highly correlated
high_corr_pairs <- which(abs(corr_matrix) > 0.8 & lower.tri(corr_matrix), arr.ind = TRUE)

high_corr_var_names <- data.frame(
  var1 = rownames(corr_matrix)[high_corr_pairs[, 1]],
  var2 = colnames(corr_matrix)[high_corr_pairs[, 2]],
  correlation = corr_matrix[high_corr_pairs]
) %>%
  arrange(var1)

high_corr_vars <- unique(c(
  rownames(corr_matrix)[high_corr_pairs[, 1]],
  colnames(corr_matrix)[high_corr_pairs[, 2]]
))

high_corr_matrix <- corr_matrix[high_corr_vars, high_corr_vars]

high_corr_var_names
```

### Produce correlation heatmap
```{r}
corrplot(high_corr_matrix,
         method = "circle",
         type = "upper",
         order = "hclust",
         tl.col = "black",
         tl.cex = 0.7,
         col = colorRampPalette(c("#005AB5", "white", "#b30404"))(200),
         title = "Predictor Correlation Heatmap (r > 0.8)",
         mar = c(0, 0, 2, 0))
```

### Addressing Redundant/Unneeded Variables
#### Table illustrating redundancy
```{r}
redundant <- high_corr_var_names %>%
  filter((var1 == "heart_disease_deaths" & var2 == "cardiovascular_disease_deaths") |
           (var1 == "opioid_overdose_deaths" & var2 == "drug_overdose_deaths") |
           (var1 == "poverty_in_children" & var2 == "poverty_in_all_ages") |
           (var1 == "renters_vs_owners" & var2 == "owner_occupied_housing") |
           (var1 == "homicides" & var2 == "gun_deaths_firearms")) %>%
  mutate(correlation = round(correlation, 2))

redundant %>%
  kbl(caption = "Redundant Variable Correlations") %>%
  kable_classic(full_width = FALSE, html_font = "Cambria")
```

#### High VIF Table
```{r}
lm <- lm(life_expectancy ~ ., data = bchc_imputed)
vif <- vif(lm)
high_vif <- data.frame(
  feature = rownames(vif),
  adjusted_gvif = round(vif[, "GVIF^(1/(2*Df))"], 2)
) %>%
  filter(adjusted_gvif > 10) %>%
  arrange(desc(adjusted_gvif))

high_vif %>%
  kbl(caption = "Features with Adjusted GVIF > 10", row.names = FALSE) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria")
```

#### Drop Redundant Variables
```{r}
vars_to_drop <- c("deaths_from_all_causes", "premature_death", "heart_disease_deaths",
                  "opioid_overdose_deaths", "poverty_in_children", "owner_occupied_housing")
bchc_final <- bchc_imputed %>% dplyr::select(-all_of(vars_to_drop))
```

#### Feature Engineering - Homicide/Gun Index
```{r}
bchc_final <- bchc_final %>%
  mutate(homicides_scaled = scale(homicides),
         gun_deaths_scaled = scale(gun_deaths_firearms),
         homicide_gun_index = rowMeans(cbind(homicides_scaled, gun_deaths_scaled)))

bchc_final <- bchc_final %>% dplyr::select(-all_of(c("homicides", "homicides_scaled", "gun_deaths_firearms", "gun_deaths_scaled")))
```

## Descriptive Statistics
### Predictor Distribution Examples
```{r}
hist_drug <- ggplot(bchc_final, aes(x = drug_overdose_deaths)) +
  geom_histogram(fill = "#b35707", color = "#c9c9c9", binwidth = 5) +
  labs(title = "Histogram of Drug Overdose Deaths",
       x = "Drug Overdose Deaths\n(Per 100,000, Age-Adjusted)",
       y = "Count of Rows") +
  hist_theme

hist_cancer <- ggplot(bchc_final, aes(x = all_cancer_deaths)) +
  geom_histogram(fill = "#470a70", color = "#c9c9c9", binwidth = 5) +
  labs(title = "Histogram of Cancer Deaths",
       x = "All Cancer Deaths\n(Per 100,000, Age-Adjusted)",
       y = "Count of Rows") +
  hist_theme

grid.arrange(hist_drug, hist_cancer, nrow=2)
```

### Life Expectancy Summary Statistics Table
```{r}
le_descriptives <- as.data.frame(unclass(summary(bchc_final$life_expectancy)))
colnames(le_descriptives) <- "Statistic Value"

le_descriptives %>%
  kbl(caption = "Descriptive Statistics: Life Expectancy") %>%
  kable_classic(full_width = FALSE, html_font = "Cambria")
```

### Life Expectancy Histogram
```{r}
# Get the highest and lowest life expectancy values and labels
max_le <- max(le_descriptives$`Statistic Value`)
max_le_row <- bchc_final %>% filter(life_expectancy == max_le)

min_le <- min(le_descriptives$`Statistic Value`)
min_le_row <- bchc_final %>% filter(life_expectancy == min_le)

max_label <- paste(
  paste0("Max Life Expectancy: ", round(max_le, 2), " Years"),
  paste("City:", max_le_row$city),
  paste("Year:", max_le_row$year),
  paste("Race:", max_le_row$race),
  paste("Sex:", max_le_row$sex),
  sep = "\n"
)

min_label <- paste(
  paste0("Min Life Expectancy: ", round(min_le, 2), " Years"),
  paste("City:", min_le_row$city),
  paste("Year:", min_le_row$year),
  paste("Race:", min_le_row$race),
  paste("Sex:", min_le_row$sex),
  sep = "\n"
)

# Plot histogram and add min/max annotations
ggplot(bchc_final, aes(x = life_expectancy)) +
  geom_histogram(fill = "#005AB5", color = "#c9c9c9", binwidth = 1) +
  annotate("text", x = 90, y = 400, label = max_label, color = "#1c6b1b", size = 3.5, fontface = "bold") +
  annotate("text", x = 65, y = 400, label = min_label, color = "#b30404", size = 3.5, fontface = "bold") +
  labs(title = "Histogram of Life Expectancy",
       x = "Life Expectancy (Years)",
       y = "Count of Rows") +
  hist_theme
```

## Potential Strong Predictors
### Cardiovascular Disease Deaths
```{r}
# Pearson Correlation Test
le_cv_corr <- cor.test(bchc_final$life_expectancy, bchc_final$cardiovascular_disease_deaths, method="pearson")
le_cv_coef <- le_cv_corr$estimate
le_cv_pval <- le_cv_corr$p.value

# Scatterplot
bchc_le_cv <- bchc_final %>%
  filter(race != "All") %>%
  mutate(race = if_else(race == "Asian", "Asian/PI", race))

ggplot(bchc_le_cv, aes(x = life_expectancy, 
                       y = cardiovascular_disease_deaths, 
                       color = race)) +
  geom_point(alpha = 0.7, size = 2.5) +
  labs(
    title = "Life Expectancy vs. Cardiovascular Disease Deaths",
    subtitle = paste0("Pearson Correlation Coefficient = ", round(le_cv_coef, 2), " (P-Value: ", le_cv_pval, ")"),
    x = "Life Expectancy (Years)",
    y = "Cardiovascular Disease Deaths\n(Per 100,000, Age-Adjusted)",
    color = "Race"
  ) +
  hist_theme +
  theme(
    legend.position = c(0.65, 0.98),
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = "white", color="black")
  )
```

### Drug Overdose Deaths
```{r,warning = FALSE}
# Spearman Correlation Test
le_do_corr <- cor.test(bchc_final$life_expectancy, bchc_final$drug_overdose_deaths, method="spearman")
le_do_coef <- le_do_corr$estimate
le_do_pval <- le_do_corr$p.value

bchc_le_drug <- bchc_final %>%
  filter(sex != "Both")

ggplot(bchc_le_drug, aes(x = life_expectancy, 
                         y = drug_overdose_deaths, 
                         color = sex)) +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_color_manual(values = c("Female" = "#b30404", "Male" = "#005AB5")) +
  labs(
    title = "Life Expectancy vs. Drug Overdose Deaths",
    subtitle = paste0("Spearman Correlation Coefficient = ", round(le_do_coef, 2), " (P-Value: ", le_do_pval, ")"),
    x = "Life Expectancy (Years)",
    y = "Drug Overdose Deaths\n(Per 100,000, Age-Adjusted)",
    color = "Sex"
  ) +
  hist_theme +
  theme(
    legend.position = c(0.85, 0.98),
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = "white", color="black")
  )
```

# MODEL BUILDING AND EVALUATION
In this section of code, I train LASSO, XGBoost, and ANN models using 10-fold single cross validation. I also perform 5-fold double cross validation. I then select the final model and evaluate performance based on MAE, R-squared, and analysis of high error points. Finally, I evaluate feature importance through importance plots, SHAP values, and partial dependence plots.


## Model Training
### Naive Model
```{r}
mean_le <- mean(bchc_final$life_expectancy)
naive_mae <- round(mean(abs(bchc_final$life_expectancy - mean_le)), 3)
print(paste0("Naive MAE: ", naive_mae))
```

### Baseline Model - LASSO
#### Display Hyperparameter Values
```{r}
data.frame(Hyperparameter = c("$\\alpha$", "$\\lambda$"),
           "Value(s)" = c("1", "100 values equally spaced from 0.0001 to 10"),
           check.names = FALSE) %>%
  kbl(caption = "LASSO Hyperparameters", booktabs = TRUE) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  row_spec(0, bold = TRUE, extra_css = "border-top: 2px solid black;")
```

#### Fit models (10-fold CV)
```{r, warning = FALSE, message = FALSE}
set.seed(10) # Set seed for reproducibility
train_method <- trainControl(method = "cv", number = 10)

fit_lasso <- train(life_expectancy ~ .,
                   data = bchc_final[, !(names(bchc_final) %in% c("city", "year", "race", "sex"))],
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 1,
                                          lambda = 10^seq(-4, 1, length=100)), # Tuning this hyperparameter
                   trControl = train_method,
                   preProcess = c("YeoJohnson", "center", "scale")) # Centered, scaled, and transformed
```

#### Display the results of the "best" LASSO model
```{r}
best_lasso <- fit_lasso$results[fit_lasso$results$MAE == min(fit_lasso$results$MAE), c(1:5)]
best_lasso
```

### XGBoost
#### Display Hyperparameter Values
```{r}
data.frame(Hyperparameter = c("Trees", "Max Tree Depth", "$\\eta$", "$\\Gamma$", "Fraction of Predictors Per Tree", "Minimum Child Weight", "Fraction of Rows Per Tree"),
           "Value(s)" = c("75, 100, or 125", "3, 4, 5, 6, or 7", "0.3", "0", "0.8", "4", "1"),
           check.names = FALSE) %>%
  kbl(caption = "XGBoost Hyperparameters", booktabs = TRUE) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  row_spec(0, bold = TRUE, extra_css = "border-top: 2px solid black;")
```

#### Fit models (10-fold CV)
```{r, warning = FALSE, message = FALSE}
set.seed(10) # Set seed for reproducibility
train_method <- trainControl(method = "cv", number = 10)

fit_xgb <- train(life_expectancy ~ ., 
              data = bchc_final[, !(names(bchc_final) %in% c("city", "year", "race", "sex"))],
              method = "xgbTree",
              tuneGrid = expand.grid(nrounds = c(75,100,125), # Tuning this hyperparameter
                                     max_depth = 3:7, # Tuning this hyperparameter
                                     eta = 0.3,
                                     gamma = 0,
                                     colsample_bytree = 0.8,
                                     min_child_weight = 4,
                                     subsample = 1),
              verbosity = 0,
              trControl = train_method)
```

#### Display the results of the "best" XGBoost model
```{r}
best_xgb <- fit_xgb$results[fit_xgb$results$MAE == min(fit_xgb$results$MAE), c(2, 7:10)]
best_xgb
```

### ANN
#### Display Hyperparameter Values
```{r}
data.frame(Hyperparameter = c("Size", "Decay"),
           "Value(s)" = c("3, 5, or 7", "0, 0.01, or 0.1"),
           check.names = FALSE) %>%
  kbl(caption = "ANN Hyperparameters", booktabs = TRUE) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  row_spec(0, bold = TRUE, extra_css = "border-top: 2px solid black;")
```

#### Before fitting ANN model, choose the best features based on LASSO and XGBoost models
```{r}
# First, find top 20 most important variables in the XGBoost model
xgb_importance <- varImp(fit_xgb)$importance
xgb_importance$Feature <- rownames(xgb_importance)

top_n <- 20
top_xgb_features <- xgb_importance %>%
  arrange(desc(Overall)) %>%
  slice_head(n = top_n) %>%
  pull(Feature)

# Second, find features with non-zero coefficients in the LASSO model
coef_lasso <- coef(fit_lasso$finalModel, s = fit_lasso$bestTune$lambda)
top_lasso_features <- rownames(coef_lasso)[which(coef_lasso != 0)][-1]


# Find the intersection of the best vars from the LASSO and XGBoost models
selected_vars <- intersect(top_xgb_features, top_lasso_features)
```

#### Fit models (10-fold CV)
```{r, warning = FALSE, message = FALSE}
set.seed(10) # Set seed for reproducibility
train_method <- trainControl(method = "cv", number = 10)

fit_ann <- train(life_expectancy ~ ., 
              data = bchc_final[, c("life_expectancy", selected_vars)],
              method = "nnet",
              tuneGrid = expand.grid(size = c(3, 5, 7), # Tuning this hyperparameter
                                     decay = c(0, 0.01, 0.1)), # Tuning this hyperparameter
              linout = TRUE,
              trace = FALSE,
              maxit = 200, # Control number of iterations
              trControl = train_method,
              preProcess = c("YeoJohnson", "range")) # Transformation + min-max scaling
```

#### Display the results of the "best" ANN model
```{r}
best_ann <- fit_ann$results[fit_ann$results$MAE == min(fit_ann$results$MAE),]
best_ann
```

### 5-Fold Double CV of entire model selection process
```{r, warning = FALSE, message = FALSE}
set.seed(10) # Set seed for reproducibility

# Setup
n <- nrow(bchc_final)
pred_los_outer <- rep(NA, n)

nfolds_outer <- 5
groups <- rep(1:nfolds_outer, length = n)
cvgroups <- sample(groups, n)

train_method <- trainControl(method = "cv", number = 10)

# Initiate outer loop
for (j in 1:nfolds_outer) {
  in_train <- (cvgroups != j)
  in_valid <- (cvgroups == j) 
  traindata.out <- bchc_final[in_train,]
  validdata.out <- bchc_final[in_valid,]
  
  # LASSO single CV
  lasso <- train(life_expectancy ~ .,
                   data = traindata.out[, !(names(traindata.out) %in% c("city", "year", "race", "sex"))],
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 1,
                                          lambda = 10^seq(-4, 1, length=100)), # Tuning this hyperparameter
                   trControl = train_method,
                   preProcess = c("YeoJohnson", "center", "scale")) # Centered, scaled, and transformed
  
  # XGBoost single CV
  xgb <- train(life_expectancy ~ ., 
              data = traindata.out[, !(names(traindata.out) %in% c("city", "year", "race", "sex"))],
              method = "xgbTree",
              tuneGrid = expand.grid(nrounds = c(75,100,125), # Tuning this hyperparameter
                                     max_depth = 3:7, # Tuning this hyperparameter
                                     eta = 0.3,
                                     gamma = 0,
                                     colsample_bytree = 0.8,
                                     min_child_weight = 4,
                                     subsample = 1),
              verbosity = 0,
              trControl = train_method)
  
  # ANN single CV
  ann <- train(life_expectancy ~ ., 
              data = traindata.out[, c("life_expectancy", selected_vars)],
              method = "nnet",
              tuneGrid = expand.grid(size = c(3, 5, 7), # Tuning this hyperparameter
                                     decay = c(0, 0.01, 0.1)), # Tuning this hyperparameter
              linout = TRUE,
              trace = FALSE,
              maxit = 200, # Control number of iterations
              trControl = train_method,
              preProcess = c("YeoJohnson", "range")) # Transformation + min-max scaling
  
  
  # Find lowest MAE
  all_MAE_CV <- c(min(lasso$results$MAE),
                  min(xgb$results$MAE),
                  min(ann$results$MAE))
  
  all_train_output <- list(lasso, xgb, ann)
  
  allbest <- which(all_MAE_CV == min(all_MAE_CV))
  whichbest <- ifelse(length(allbest) == 1, allbest, sample(allbest,1))
  bestmodel_train_output <- all_train_output[[whichbest]]

  pred_los <- bestmodel_train_output %>% predict(validdata.out)
  pred_los_outer[in_valid] <- pred_los
}

# Display performance metrics
mae <- round(mean(abs(pred_los_outer - bchc_final$life_expectancy)), 3)
r2 <- round(1 - sum((bchc_final$life_expectancy - pred_los_outer)^2) / sum((bchc_final$life_expectancy - mean(bchc_final$life_expectancy))^2), 3)
sprintf("MAE: %s", mae)
sprintf("R-squared: %s", r2)
```

### Display performance metrics
```{r, purl = FALSE}
lasso_mae <- round(best_lasso$MAE, 3)
lasso_r2 <- round(best_lasso$Rsquared, 3)
xgb_mae <- round(best_xgb$MAE, 3)
xgb_r2 <- round(best_xgb$Rsquared,3)
ann_mae <- round(best_ann$MAE, 3)
ann_r2 <- round(best_ann$Rsquared, 3)
lasso_hyperparams = paste0("$\\lambda$", " = ", round(best_lasso$lambda, 5))
xgb_hyperparams = paste0("Trees = ", best_xgb$nrounds, "; Max Tree Depth = ", best_xgb$max_depth)
ann_hyperparams = paste0("Size = ", best_ann$size, "; Decay = ", best_ann$decay)

data.frame(Model = c("LASSO", "XGBoost", "ANN"),
           MAE = c(lasso_mae, xgb_mae, ann_mae),
           `R²` = c(lasso_r2, xgb_r2, ann_r2),
           "Tuned Hyperparameters" = c(lasso_hyperparams, xgb_hyperparams, ann_hyperparams),
           check.names = FALSE) %>%
  kbl(caption = "Model Performance Metrics", booktabs = TRUE) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  row_spec(0, bold = TRUE, extra_css = "border-top: 2px solid black;")
```

### Select final model and add predictions to dataframe
```{r, warning = FALSE, message = FALSE}
best_model <- fit_xgb$finalModel
X <- as.matrix(bchc_final[, names(fit_xgb$trainingData)[-1]])
bchc_final$pred_life_expectancy <- predict(best_model, X)
bchc_final$residual <- bchc_final$life_expectancy - bchc_final$pred_life_expectancy
```

## Model Evaluation
### Histogram of Residuals
```{r}
ggplot(bchc_final, aes(x = residual)) +
  geom_histogram(fill = "#005AB5", color = "#c9c9c9", binwidth = 0.25) +
  geom_vline(xintercept=-2.5, linetype = "dashed", color="#b30404", size=1) +
  geom_vline(xintercept=2.5, linetype = "dashed", color="#b30404", size=1) +
  labs(title = "Histogram of Model Residuals",
       x = "Residual (Observed - Predicted Life Expectancy)",
       y = "Count of Rows") +
  hist_theme
```

### Scatterplot of Actual Life Expectancy vs. Residuals
```{r}
ggplot(bchc_final, aes(x = life_expectancy, 
                       y = residual,
                       color = abs(residual) > 2.5)) +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_color_manual(
    values = c("FALSE" = "#005AB5", "TRUE" = "#b30404"),
    labels = c("<= 2.5", "> 2.5"),
    name = "|Residual|"
  ) +
  labs(
    title = "Life Expectancy vs. Residuals",
    x = "Actual Life Expectancy (Years)",
    y = "Residuals\n(Observed - Predicted Life Expectancy)"
  ) +
  hist_theme +
  theme(
    legend.position = c(0.8, 0.35),
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = "white", color="black")
  )
```

### Frequency of High Error Data Points by Race/Sex
```{r}
high_error <- bchc_final[abs(bchc_final$residual) > 2.5, ]

race_sex_freq <- high_error %>%
  count(race, sex) %>%
  pivot_wider(names_from = sex, values_from = n, values_fill = 0)

race_sex_freq <- race_sex_freq %>%
  rowwise() %>%
  mutate(Total = sum(c_across(where(is.numeric)))) %>%
  rename(Race = race) %>%
  arrange(desc(Total))

total_row <- data.frame(Race = "Total",
                        Both = sum(race_sex_freq$Both),
                        Female = sum(race_sex_freq$Female),
                        Male = sum(race_sex_freq$Male),
                        Total = sum(race_sex_freq$Total))

race_sex_freq <- bind_rows(race_sex_freq, total_row)

race_sex_freq %>%
  kbl(caption = "Race and Sex Frequency for High Errors", booktabs = TRUE) %>%
  add_header_above(c(" " = 1, "Sex" = 3, " " = 1)) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  row_spec(0, bold = TRUE, extra_css = "border-top: 2px solid black;") %>%
  row_spec(nrow(race_sex_freq), bold = TRUE, extra_css = "border-top: 2px solid black;")
```

## Boxplots of Asian/PI and Hispanic Populations of Both Sexes Actual and Predicted Life Expectancy
```{r}
# Filter Data
boxplot_data <- bchc_final %>% filter(race %in% c("Asian/PI", "Hispanic"), sex == "Both")
boxplot_outliers_actual <-count_outliers(boxplot_data[["life_expectancy"]])
boxplot_outliers_pred <-count_outliers(boxplot_data[["pred_life_expectancy"]])

# Actual Life Expectancy Boxplot
ggplot(boxplot_data, aes(x = reorder(race, life_expectancy, FUN = median), y = life_expectancy,  fill = race)) +
  geom_boxplot(
    outlier.colour = "#b30404",      
    outlier.shape = 8,
    outlier.size = 2
  ) +
  annotate("text", x = 1.65, y = 70, 
           label = paste("Outliers:", boxplot_outliers_actual),
           hjust = 1.1, vjust = -0.5, color = "#b30404", size = 4) +
  scale_fill_manual(values = c("Hispanic" = "#2ecc71", "Asian/PI" = "#3498db")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Boxplot of Life Expectancy by Race",
       subtitle = "Asian/PI and Hispanic Populations of Both Sexes",
       x = "Race",
       y = "Life Expectancy") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18, color = "#2c3e50", hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 14, color = "#34495e", hjust = 0.5),
    axis.title.y = element_text(size = 14, face = "bold", color = "#2c3e50"),
    axis.text.y = element_text(size = 12, color = "#2c3e50"),
    axis.title.x = element_text(size = 14, face = "bold", color = "#2c3e50"),
    axis.text.x = element_text(size = 12, color = "#2c3e50"),
    legend.position = "none"
  )

# Predicted Life Expectancy Boxplot
ggplot(boxplot_data, aes(x = reorder(race, pred_life_expectancy, FUN = median), y = pred_life_expectancy,  fill = race)) +
  geom_boxplot(
    outlier.colour = "#b30404",      
    outlier.shape = 8,
    outlier.size = 2
  ) +
  annotate("text", x = 1.65, y = 70, 
           label = paste("Outliers:", boxplot_outliers_pred),
           hjust = 1.1, vjust = -0.5, color = "#b30404", size = 4) +
  scale_fill_manual(values = c("Hispanic" = "#2ecc71", "Asian/PI" = "#3498db")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Boxplot of Predicted Life Expectancy by Race",
       subtitle = "Asian/PI and Hispanic Populations of Both Sexes",
       x = "Race",
       y = "Predicted Life Expectancy") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18, color = "#2c3e50", hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 14, color = "#34495e", hjust = 0.5),
    axis.title.y = element_text(size = 14, face = "bold", color = "#2c3e50"),
    axis.text.y = element_text(size = 12, color = "#2c3e50"),
    axis.title.x = element_text(size = 14, face = "bold", color = "#2c3e50"),
    axis.text.x = element_text(size = 12, color = "#2c3e50"),
    legend.position = "none"
  )
```

### SHAP
#### Calculate SHAP Values
```{r}
shap_values <- shap.values(xgb_model = best_model, X_train = X)
shap_long <- shap.prep(xgb_model = best_model, X_train = X)
```

#### Display largest over and under-predictions
```{r}
highest_error <- high_error %>%
  arrange(desc(residual)) %>%
  mutate(life_expectancy = round(life_expectancy, 3),
         pred_life_expectancy = round(pred_life_expectancy, 3),
         residual = round(residual, 3)) %>%
  head(1)

lowest_error <- high_error %>%
  arrange(residual) %>%
  mutate(life_expectancy = round(life_expectancy, 3),
         pred_life_expectancy = round(pred_life_expectancy, 3),
         residual = round(residual, 3)) %>%
  head(1)

bind_rows(highest_error, lowest_error) %>%
  dplyr::select(c(city, year, race, sex, life_expectancy, pred_life_expectancy, residual)) %>%
  rename("City" = city,
         "Year" = year,
         "Race" = race,
         "Sex" = sex,
         "Actual Life Expectancy" = life_expectancy,
         "Predicted Life Expectancy" = pred_life_expectancy,
         "Residual" = residual) %>%
  kbl(caption = "Largest Over-Prediction and Under-Prediction", booktabs = TRUE) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  row_spec(0, bold = TRUE, extra_css = "border-top: 2px solid black;")
```

#### SHAP plots for largest over and under-predictions
```{r}
# Find largest residual points and corresponding SHAP values
highest_error_preds <- highest_error %>% dplyr::select(-c(city, year, race, sex, life_expectancy, pred_life_expectancy, residual)) %>% as.matrix()
lowest_error_preds <- lowest_error %>% dplyr::select(-c(city, year, race, sex, life_expectancy, pred_life_expectancy, residual)) %>% as.matrix()
highest_idx <- which(apply(X, 1, function(y) all(y == highest_error_preds[1, ])))
lowest_idx <- which(apply(X, 1, function(y) all(y == lowest_error_preds[1, ])))
shap_highest <- as.numeric(shap_values$shap_score[highest_idx, ])
shap_lowest <- as.numeric(shap_values$shap_score[lowest_idx, ])

highest_df <- data.frame(
  Feature = colnames(X),
  SHAP = shap_highest
)

lowest_df <- data.frame(
  Feature = colnames(X),
  SHAP = shap_lowest
)

# Plot SHAP values
ggplot(highest_df, aes(x = reorder(Feature, SHAP), y = SHAP, fill = SHAP > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "SHAP for Largest Under-Prediction",
       subtitle = "Hispanic Males Milwaukee, WI in 2019",
       y = "SHAP Value",
       x = "") +
  hist_theme +
  theme(legend.position = "none")

ggplot(lowest_df, aes(x = reorder(Feature, SHAP), y = SHAP, fill = SHAP > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "SHAP for Largest Over-Prediction",
       subtitle = "Hispanics of Both Sexes Portland, OR in 2020",
       y = "SHAP Value",
       x = "") +
  hist_theme +
  theme(legend.position = "none")
```

## Feature Importance
### Feature Importance Plot
```{r, warning = FALSE, message = FALSE}
imp <- xgb.importance(feature_names = best_model$feature_names,
               model = best_model)

xgb.ggplot.importance(imp, top_n = 15) +
  ggtitle("Top 15 Most Important Features in Final Model") +
  hist_theme
```

### Overall SHAP Plot
```{r}
shap.plot.summary(shap_long)
```

### Partial Dependency Plots
```{r}
# Select example predictors for PDPs
pdp_vars <- c(
  "injury_deaths",
  "cardiovascular_disease_deaths",
  "all_cancer_deaths",
  "diabetes_deaths",
  "per.capita_household_income",
  "uninsured_all_ages"
)
pdp_plots <- list()

# Generate a plot for each predictor
for (v in pdp_vars) {
  
  pdp_df <- partial(
    object = best_model,
    pred.var = v,
    train = as.data.frame(X),
  )
  
  p <- ggplot(pdp_df, aes_string(x = v, y = "yhat")) +
    geom_line(color = "#2E86AB", size = 1.2) +
    labs(
      title = v,
      x = "Predictor Value",
      y = "Predicted Life Expectancy"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 9, color = "#2c3e50"),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
    )
  
  pdp_plots[[v]] <- p
}

# Arrange plots in grid
grid.arrange(pdp_plots[[1]], pdp_plots[[2]], pdp_plots[[3]], pdp_plots[[4]], pdp_plots[[5]], pdp_plots[[6]], nrow=2, ncol=3,
             bottom = textGrob("Predictor Value", gp = gpar(fontsize = 14)),
             left = textGrob("Average Predicted Life Expectancy", rot = 90, gp = gpar(fontsize = 14)))
```

# RESULTS, INSIGHTS, AND RECOMMENDATIONS
In this section of code, I determine three predictors having a large negative impact on life expectancy for the most vulnerable demographic groups in Long Beach, CA. I then inspect some overall trends and statistics related to these factors to support the narrative in my final paper.

## Long Beach, CA - 2023
### Find the five lowest life expectancy demographic groups
```{r}
lb2023_lowest_df <- bchc_final %>%
  filter(city == "Long Beach, CA",
         year == "2023",
         race != "All",
         sex != "Both") %>%
  arrange(pred_life_expectancy) %>%
  head(5)

lb2023_lowest_df %>%
  dplyr::select(c(race, sex, pred_life_expectancy)) %>%
  rename(Race = race,
         Sex = sex,
         `Predicted Life Expectancy` = pred_life_expectancy) %>%
  kbl(caption = "Five Demographic Groups with Lowest Predicted Life Expectancy; Long Beach, CA; 2023", booktabs = TRUE) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  row_spec(0, bold = TRUE, extra_css = "border-top: 2px solid black;")
```

### Inspect SHAP values for Lowest Predicted Life Expectancy Populations in Long Beach, CA in 2023
```{r}
# Find features with negative SHAP values for our case study
lb2023_lowest <- lb2023_lowest_df %>%
  dplyr::select(-c(city, year, race, sex, life_expectancy, pred_life_expectancy, residual)) %>%
  as.matrix()

lb2023_lowest_idx <- which(apply(X, 1, function(y) all(y == lb2023_lowest[1, ])))
shap_lb2023_lowest <- as.numeric(shap_values$shap_score[lb2023_lowest_idx, ])

lb2023_lowest_shap <- data.frame(
  Feature = colnames(X),
  SHAP = shap_lb2023_lowest
) %>%
  filter(SHAP < 0)

# Plot SHAP values
ggplot(lb2023_lowest_shap, aes(x = reorder(Feature, SHAP), y = SHAP, fill = SHAP > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Negative SHAP Values for Long Beach in 2023",
       subtitle = "Limited to the Five Demographic Groups\nwith the Lowest Predicted Life Expectancy",
       y = "SHAP Value",
       x = "") +
  hist_theme +
  theme(legend.position = "none")
```

### Inspect Cardiovascular Disease Deaths by City in 2023
```{r}
bchc_final %>%
  filter(year == "2023",
         race == "All",
         sex == "Both") %>%
  dplyr::select(c(city, cardiovascular_disease_deaths)) %>%
  arrange(desc(cardiovascular_disease_deaths))
```

### Trend Motor Vehicle Deaths Over Time
```{r}
lb_motor <- bchc_final %>%
  filter(city == "Long Beach, CA",
         race == "All",
         sex == "Both") %>%
  dplyr::select(year, motor_vehicle_deaths) %>%
  mutate(year = as.numeric(as.character(year)))

ggplot(lb_motor, aes(x = year, y = motor_vehicle_deaths)) +
  geom_line(size = 1.2) +
  geom_vline(xintercept=2020, linetype = "dashed", color="#b30404", size=1) +
  scale_x_continuous(breaks = unique(lb_motor$year)) +
  annotate(
    "text",
    x = 2020.4,
    y = 8.6,
    label = "Safe Streets Long Beach Action Plan",
    color = "#b30404",
    angle = 90,
    hjust = 0.5,
    vjust = 0
  ) +
  labs(
    title = "Motor Vehicle Deaths Per Year",
    subtitle = "All Demographic Groups; Long Beach, CA",
    x = "Year",
    y = "Motor Vehicle Deaths\n(Per 100,000, Age-Adjusted)"
  ) +
  hist_theme +
  theme(panel.grid.minor.x = element_blank())
```

### Inspect Diabetes Deaths by City in 2023
```{r}
bchc_final %>%
  filter(year == "2023",
         race == "All",
         sex == "Both") %>%
  dplyr::select(c(city, diabetes_deaths)) %>%
  arrange(desc(diabetes_deaths))
```

### Inspect Diabetes Death Rates between Demographic Groups
```{r}
lb_diabetes <- bchc_final %>%
  filter(
    city == "Long Beach, CA",
    year == "2023",
    race %in% c("Black", "White"),
    sex %in% c("Female", "Male")
  ) %>%
  dplyr::select(race, sex, diabetes_deaths)

ggplot(lb_diabetes, aes(x = race, y = diabetes_deaths, fill = race)) +
  geom_col() +
  facet_wrap(~ sex, ncol = 2) +
  labs(
    title = "2023 Diabetes Deaths in Long Beach, CA",
    subtitle = "By Race and Sex",
    x = "Race",
    y = "Diabetes Deaths\n(Per 100,000, Age-Adjusted)"
  ) +
  hist_theme +
  theme(legend.position = "none")
```